---
title: "gllvm_benthos"
output: html_document
date: "2025-12-11"
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

{
  
  library(dplyr)
  
  library(INLA)
  
  library(gllvm)
  
  library(RANN)
  
  library(lattice)
  
  library(ggplot2)
  
  library(data.table)
  
  library(fmesher)
  
  library(tidyr)
  
  library(cowplot)
  
  library(RColorBrewer)
  
  library(viridis)
  
  library(ggpubr)
  
  library(MASS)
  
  library(fields) 
  
  library(pscl)
  
  library(sp)
  
  library(stringr)
  
  library(mgcv)

  library(corrplot)
  
  library(sf)
 
}

```

# Define location of data
```{r}
# Read your data
data_loc <- "/home/storage/benthos_data"
```


# Load Benthos data

```{r}
# Read your data
sub_benthos <- fread(file.path(data_loc,"sub_benthos.csv"))

# ensure counts are numeric
sub_benthos$count <- as.numeric(sub_benthos$count)
```

# Filter the data for the Zand uit Zee survey
We only want to run the model, and later the predictions for species that are also present at the extration sites
```{r}
ZanduitZee <- sub_benthos %>% 
  filter(dataset_name == "ZanduitZee")
```

```{r}
# ignoring all entries where we only got the genus
ZanduitZee <- ZanduitZee[
  sapply(strsplit(ZanduitZee$species, " "), length) == 2, ]

# to make sure the species are well represented 
species_sandpit <- ZanduitZee %>%
  filter(count > 0) %>%                       
  group_by(species) %>%
  summarise(n_locations = n_distinct(location)) %>%
  filter(n_locations >= 30)  # species should occur at least at 30 different sites (out of 166) can still be changed 

# giving us 42 different species; species from the first project are present. So we can compare the model outputs later.
# (Abra alba, Spisula subtruncata, Ophiura ophiura, Lanice conchilega, Echinocardium cordatum)
```

```{r}
# filtering the benthos data set based on those 42 species
# filtering the data also by year, year 2010 gives us data for 13 years, less years allows for a faster model run
# with 13 years the model runs several hours (I had let it run overnight). For test run code, suggestion to reduce the number of years
sub_benthos_filtered <- sub_benthos %>%
  filter(species %in% species_sandpit$species)%>% 
  filter(year > 2022)

unique(sub_benthos_filtered$species) # and only the 42 species are remaining 
```

# harmonize the different gears
```{r}
table(sub_benthos_filtered$gear)

sub_benthos_filtered$gear <- ifelse(sub_benthos_filtered$gear %in% c("MONS_EBB_MZB_Boxcorer", "MWTL_MZB_Boxcorer", "MONEOS_WS_MON", "ZuZ_Chronoseq_Boxcorer", "MWTL_MZB_Hamon", "Boxcorer"),"boxcore", sub_benthos_filtered$gear)

sub_benthos_filtered$gear <- ifelse(sub_benthos_filtered$gear %in% c("Bodemschaaf","bottom dredge" ),"WMR dredge", sub_benthos_filtered$gear)
sub_benthos_filtered$gear <- ifelse(sub_benthos_filtered$gear %in% c("dredge","MWTL_MZB_Schaaf"),"NIOZ triple D", sub_benthos_filtered$gear)
sub_benthos_filtered$gear <- ifelse(sub_benthos_filtered$gear %in% c("Zuigkor"),"dredge", sub_benthos_filtered$gear)

sub_benthos_filtered <- subset(sub_benthos_filtered, 
                            gear != "Oesterhapper" &
                            gear  != "KIER_DELTA_2021" &
                            gear != "WOZEP_PAWP_T15" &
                            gear  != "onbekend" &
                            gear != "Van Veenhapper")

table(sub_benthos_filtered$gear)
```

# Change coordinate system of the benthos data into UTM31N
```{r env}
UTM31U <- CRS("+init=epsg:32631") 
wgs84_crs <- CRS("+init=epsg:4326")

coord_e <- st_as_sf(sub_benthos_filtered, coords = c("lon", "lat"), crs = wgs84_crs)

coords_e <- st_transform(coord_e, UTM31U)
sub_benthos_filtered <- cbind(coords_e, st_coordinates(coords_e))

sub_benthos_filtered$geometry <- NULL
```
# Load the Environmental variables
```{r}
#** Bed Shear Stress ----
shear_stress_UTM <- readRDS(file.path(data_loc,"shear_stress_UTM.rds"))

#** BPI ----
bpi_UTM <- readRDS(file.path(data_loc,"bpi_UTM_all.rds"))

# mud
mud_UTM <- readRDS(file.path(data_loc,"mud_model.rds"))
```

## Add covariates to species set
```{r asign}
#' add BPI  
closest_bpi <- nn2(bpi_UTM[,c("X","Y")], sub_benthos_filtered[,c("X","Y")], k = 1, searchtype = "radius", radius = 5000)
closest_bpi$nn.idx[closest_bpi$nn.idx==0] <- NA
sub_benthos_filtered$bpi20 <- st_drop_geometry(bpi_UTM[closest_bpi$nn.idx,"bpi20"])
sub_benthos_filtered$bpi20<- round(sub_benthos_filtered$bpi20, 2)

#' bottom shear stress
closest <- nn2(shear_stress_UTM[,c("X", "Y")], sub_benthos_filtered[,c("X","Y")], k = 1, searchtype = "radius", radius = 5000)
closest$nn.idx[closest$nn.idx==0]<-NA
sub_benthos_filtered$shear_stress <- st_drop_geometry(shear_stress_UTM[closest$nn.idx,]$stress)
sub_benthos_filtered$shear_stress <- round(sub_benthos_filtered$shear_stress, 1)

#'mud
closest <- nn2(mud_UTM[,c("X", "Y")], sub_benthos_filtered[,c("X","Y")], k = 1, searchtype = "radius", radius = 5000)
closest$nn.idx[closest$nn.idx==0]<-NA
sub_benthos_filtered$mud <- st_drop_geometry(mud_UTM[closest$nn.idx,]$prediction)
sub_benthos_filtered$mud <- round(sub_benthos_filtered$mud, 2)

```

```{r data_loss_env_vars}
# making sure to keep only locations with both benthos and environmental variables
nrow(sub_benthos_filtered)
nrow(sub_benthos_filtered[!(is.na(sub_benthos_filtered$bpi20)|is.na(sub_benthos_filtered$shear_stress)|is.na(sub_benthos_filtered$mud)),])

ggplot(sub_benthos_filtered, aes(x = X, y = Y, color = factor(!(is.na(sub_benthos_filtered$bpi20)|is.na(sub_benthos_filtered$shear_stress)|is.na(sub_benthos_filtered$mud))))) +
     geom_point() +
     scale_color_manual(values = c("blue", "red"), labels = c("Missing Values", "No Missing Values")) +
     labs(title = "Points with Missing Values vs No Missing Values", x = "Latitude", y = "Longitude", color = "Legend")

sub_benthos_filtered <- sub_benthos_filtered[!(is.na(sub_benthos_filtered$bpi20)|is.na(sub_benthos_filtered$shear_stress)|is.na(sub_benthos_filtered$mud)),]
```

# Replace outliers
```{r outlier}
cols_to_adjust <- c("bpi20", "mud", "shear_stress")
summary(sub_benthos_filtered[cols_to_adjust])

# Define a function to replace outliers with quantile values
replace_outliers <- function(x) {
  lower_bound <- quantile(x, 0.01, na.rm = TRUE)
  upper_bound <- quantile(x, 0.99, na.rm = TRUE)
  x[x < lower_bound] <- lower_bound
  x[x > upper_bound] <- upper_bound
  return(x)
}
```

# prediction data set
Before creating polynomial terms, we must combine the benthos data and the prediction data into one dataset. This is necessary because:

1) The prediction dataset must contain the same environmental variables
(e.g., bpi_1, bpi_2, etc.) as the benthos dataset. Otherwise, the model cannot make predictions.

2) Polynomials must be created on the combined dataset, not separately.
If we create poly() separately for each dataset, they are calculated using different value ranges.
This results in different polynomial scaling and leads to predictions errors.
```{r add prediciton grid}
grid <- readRDS(file.path(data_loc,"mud_pred_pit.rds")) # Prediction sand pit sites

grid <-grid[,c("X" ,"Y","depth","bpi20","stress","site", "prediction")]

colnames(grid)[colnames(grid) == "stress"] <- "shear_stress"
colnames(grid)[colnames(grid) == "prediction"] <- "mud"
```

```{r}
# filter for only the environmental variables of the benthos data set
# as grid as less columns
env <- sub_benthos_filtered[,c("bpi20","shear_stress","mud")]
env$id <- "train" # introduced to separate the two environments later
env$site <- "none"

env_pred <- grid[,c("bpi20","shear_stress","mud", "site")]
env_pred$id <- "pred" # introduced to separate the two environments later

# combine
env_total <- rbind(env, env_pred)
```

# creating polynomials
```{r smooth}
bpi_poly<-poly(env_total$bpi20,4, raw = FALSE)
colnames(bpi_poly)<-paste("bpi_",1:4,sep="")
env_total<-cbind(env_total,bpi_poly)
N<-nrow(env_total)

mud_poly<-poly(env_total$mud,4, raw = FALSE)
colnames(mud_poly)<-paste("mud_",1:4,sep="")
env_total<-cbind(env_total,mud_poly)
N<-nrow(env_total)

stress_poly<-poly(env_total$shear_stress,4, raw = FALSE)
colnames(stress_poly)<-paste("stress_",1:4,sep="")
env_total<-cbind(env_total,stress_poly)
N<-nrow(env_total)

```

```{r}
env_pred  <- env_total[env_total$id == "pred", ]
env_benthos  <- env_total[env_total$id == "train", ]

env_benthos <- env_benthos[,c("bpi_1","bpi_2","bpi_3","bpi_4","mud_1","mud_2","mud_3","mud_4", "stress_1","stress_2","stress_3","stress_4")]
```

# add polynomials to benthos data

```{r}
sub_benthos_filtered <- cbind(sub_benthos_filtered, env_benthos)
```

# keep only species for which we have at least a single positive count
```{r}
spec_num <- aggregate(count~ species, data = sub_benthos_filtered, FUN = "sum")
spec_to_keep <- spec_num[spec_num$count >0,"species"]

sub_benthos_filtered <- sub_benthos_filtered[sub_benthos_filtered$species %in% spec_to_keep,]

```


# create factors
```{r}
sub_benthos_filtered <- within (sub_benthos_filtered,{
                      lsurfacearea <- log(surfacearea)
                      fYear <- factor(year)
                      fGear <- factor(gear, levels=c("boxcore", "WMR dredge", "NIOZ triple D","dredge"))
                      })
```


# changing data format

The gllvm package requires data in a site × species matrix format:

Each row represents one sampling location (or site).
Each column represents one species.The values in the matrix are the species counts.
```{r}
sub_benthos_wide <- sub_benthos_filtered %>%
    group_by(dataset_name, X,Y, fYear, fGear, lsurfacearea, species, bpi20, mud, shear_stress,bpi_1, bpi_2,bpi_3,bpi_4, stress_1,stress_2,stress_3,stress_4, mud_1,mud_2,mud_3,mud_4) %>%
    summarise(count = sum(count), .groups = "drop") %>%
    pivot_wider(
        names_from = species,
        values_from = count,
        values_fill = list(count = 0)
    )
```



# preparing for gllvm model

The GLLVM model requires two separate inputs: 
1) a species abundance matrix with one column per species and one row per site
2) a corresponding environmental dataset containing environmental predictors for the same sites
```{r}
species_cols <- setdiff( # includes all the columns that are not species
  colnames(sub_benthos_wide),
  c("lon", "lat","X","Y", "unique_ID", "fGear","lsurfacearea", "location", "dataset_name", "fYear", "mud", "shear_stress", "bpi20","bpi_1", "bpi_2","bpi_3","bpi_4", "stress_1","stress_2","stress_3","stress_4", "mud_1","mud_2","mud_3","mud_4")
)

Yspecies <- sub_benthos_wide %>%
  dplyr::select(all_of(species_cols)) %>%
  as.matrix()

Xenv <- sub_benthos_wide %>%
  dplyr::select(fGear, fYear, mud, shear_stress, bpi20,bpi_1, bpi_2,bpi_3,bpi_4, stress_1,stress_2,stress_3,stress_4, mud_1,mud_2,mud_3,mud_4)

# our offset
lsurfacearea <- sub_benthos_wide$lsurfacearea
```

# creating spatial dependency

The spatial GLLVM requires two ingredients:
1) A site ID for each unique sampling location
2) A distance (or coordinates) matrix describing the spatial relationships between sites

The latent variables in a spatial GLLVM are allowed to be spatially correlated.
This means the model can capture spatial structure in community composition
that is not explained by the environmental covariates.

However, so far I did not manage to make this work. It always ended in R studio crash as it took up too much memory.
I suspect: 
- issue with repeated sample locations
- issue with the amount of locations

Paper to read: https://doi.org/10.1111/biom.13416
```{r}
# Create site ID based on X and Y coordinates
# sub_benthos_wide$siteID <- seq_len(nrow(sub_benthos_wide))

# sub_benthos_wide <- within(sub_benthos_wide,{
#                                 fsiteID <- factor(siteID)
#                       })
# 
# site <- factor(sub_benthos_wide$siteID)
# coords <- cbind(sub_benthos_wide$X, sub_benthos_wide$Y)
# rownames(coords) <- levels(site)

# Spatial GLLVMs use a correlation structure based on distances between sites.
# dist() computes Euclidean distances between all pairs of sites.

# site_dist <- as.matrix(dist(coords))
```

# defining the study design
studyDesign argument is used to specify the experimental or observational structure of the data, particularly when there are random effects or repeated measures.

It can be used as:

- Grouping factors for random effects:
  tells the model which variables define the grouping of sites or samples, e.g. if you have repeated sampling in different years (fYear)

- Controlling for hierarchical structure
  Sites sampled in the same year may be more similar to each other than to other years

```{r}
sub_benthos_wide$siteID <- interaction(sub_benthos_wide$X, sub_benthos_wide$Y, drop = TRUE)

# Convert siteID to consecutive numeric IDs (1, 2, 3, ...)
sub_benthos_wide$site <- as.numeric(factor(sub_benthos_wide$siteID))

studyDesign = data.frame(fgear = factor(sub_benthos_wide$fGear),
                         fyear = factor(sub_benthos_wide$fYear),
                         site  = factor(sub_benthos_wide$site)
                         )
```



# model
Warnings about zeros in the species data set are expected ()
```{r}
model_1 <- gllvm(
  y = Yspecies,
  X = Xenv,
  formula = ~ fGear,
  studyDesign = studyDesign, 
  row.eff = ~ corAR1(1|fyear), # works well with multiple years (e.g. 13 years), with fewer years fYear in the model formula
  offset = lsurfacearea,
  starting.val = "res",  #starting values can be generated by fitting model without latent variables, 
                         # and applying factorial analysis to residuals to get starting values
                         # for latent variables and their coefficients (starting.val = "res"). This approach is default and recommended
  family = "negative.binomial",
  method = "VA"         # model can be fitted using Laplace approximation method (method = "LA") or 
                        # variational approximation method (method = "VA"), or with extended variational
                        # approximation method (method = "EVA") when VA is not applicable. If particular model has not been implemented using the selected method, model is fitted
                        # using the alternative method as a default. Defaults to "VA".
  
                        # predictLVs.gllvm only works with VA models. Current limitation of gllvm
)

model_2 <- gllvm(
  y = Yspecies,
  X = Xenv,
  formula = ~ fGear,
  offset = lsurfacearea,
  studyDesign = studyDesign,
  row.eff = ~ (1|site) + corAR1(1|fyear),
   starting.val = "res",
  family = "negative.binomial",
   method = "VA"
)
```

```{r}
AIC(model_1)
AIC(model_2)
```

# adding enviornmanetal variables

```{r}
model_3 <- gllvm(
  y = Yspecies,
  X = Xenv,
  formula = ~ fGear + bpi20 + shear_stress + mud,
  offset = lsurfacearea,
  studyDesign = studyDesign, 
  row.eff = ~ corAR1(1|fyear),
  starting.val = "res",
  family = "negative.binomial",
   method = "LA"
) 

model_4 <- gllvm(
  y = Yspecies,
  X = Xenv,
  formula = ~ fGear + bpi_1+ bpi_2 +bpi_3 + bpi_4 + stress_1+ stress_2+ stress_3+ stress_4 + mud_1+ mud_2+ mud_3+ mud_4,
  offset = lsurfacearea,
  studyDesign = studyDesign, 
  row.eff = ~ corAR1(1|fyear),
  starting.val = "res",
  family = "negative.binomial",
   method = "LA"
) 
```

```{r}
AIC(model_3) 
AIC(model_4)
```

# Species -species assocations
```{r}
library(corrplot); library(gclus)

cr <- getResidualCor(model_4)
corrplot(cr[order.single(cr), order.single(cr)], diag = FALSE, type = "lower", 
         method = "square", tl.cex = 0.5, tl.srt = 45, tl.col = "red")
```

```{r}
summary(model_4)
```

```{r}
# Graphical visualization of model coefficients
coefplot(model_4, cex.ylab = 0.7, mar = c(4, 9, 2, 1), mfrow=c(1,1))

# Compute confidence intervals for the environmental coefficients.
confint(model_4, level = 0.95, parm = "Xcoef")
```

```{r}
# Generic plotting function for a GLLVM object
# It shows:
# residual diagnostics
# Latent variable plots

plot(model_4, mfrow = c(3,2))
```

```{r}
# Ordination (biplot) of species and sites in the latent variable space.
ordiplot(model_4, biplot = TRUE, ind.spp = 18, xlim = c(-3, 3), ylim = c(-3, 3), 
         main = "Biplot")
```

```{r}
# Computes variance partitioning for the model
# Separates the proportion of variance explained by:
# - Environmental predictors (X)
# - Latent variables
# - Residual unexplained variance

VP1 <- VP(model_4)
plot(VP1, args.legend = list(cex=0.5), col=hcl.colors(16, "Roma"))
```

```{r}
# examining AR1 temporal random effects
model_Pit4$params$sigma

rowYear <- model_Pit4$params$row.params
predErr <- getPredictErr(model_Pit4)
seYear  <- predErr$row.effects[[1]]   # only one group
years <- as.numeric(levels(studyDesign$fyear))
plot(years, rowYear,
     ylim = range(rowYear) + c(-1.96,1.96)*max(abs(seYear)),
     main = "Random effect: year",
     xlab = "Year", ylab = "Random effect")

lines(years, rowYear - 1.96*seYear, col = 2)
lines(years, rowYear + 1.96*seYear, col = 2)
```

```{r}
# # species optima -> only if the latent variables are quadratic, so far I did not fully understood this yet
# But in theory it should allow:
# with quadratic latent variables, species responses along the latent gradients can be characterized in terms of optima and tolerances. The optima() function  
# estimates the position along each latent variable where a species is expected to reach its highest abundance, representing its preferred ecological niche. The 
# tolerances() function quantifies the niche breadth, indicating whether a species is specialized (narrow tolerance) or 
# generalist (wide tolerance) along the gradient.
# ## S3 method for class 'gllvm'
# optima(model_Pit4, sd.errors = TRUE, ...)
# ## S3 method for class 'gllvm'
# tolerances(object, sd.errors = TRUE, ...)
```

# Predict abundance for sand pit areas
```{r add prediciton grid}
# remember the environmental dataset we already created
head(env_pred)

# we do not need "id" anymore, identify as new Environment 
Xnew <- env_pred %>%
  dplyr::select(- id)
```

```{r}
# we need year and gear
new_grid <- Xnew %>%
  mutate(fYear = NA,
         fGear = NA)

new_grid <- within (new_grid,{
                      fYear <- factor(2023, levels = levels(Xenv$fYear))
                      fGear <- factor("WMR dredge", levels = levels(Xenv$fGear))
                      })

# what is the mean lsurfacearea in our dataset?
summary(sub_benthos_wide$lsurfacearea)
newOffset <- rep(2.688, nrow(new_grid))
```

```{r}
# run the predictions
# adding offset is causing errors 
# new sites can be specified in newX. If predictors newX (and newTR) are given, and newLV is not, latent
# variables are not used in the predictions. 
# level stands for latent variables: level =1 (latent variable are used, gives errors as different numbers of locations)

x <-  predict.gllvm(model_Pit4, new_grid, offset= FALSE, level = 0)

# getting the location information to be able to plot the predictions 
location <- grid %>%
  dplyr::select(X, Y, site)

prediction <- cbind(x, location)
```

# plot prediction

```{r}
# Load packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
library(RColorBrewer)
library(ggspatial)
library(patchwork)
library(ggh4x)

# --------------------------
# Select species to plot
# --------------------------

species_list <- c("Spisula subtruncata", "Abra alba", "Ophiura ophiura", "Lanice conchilega", "Echinocardium cordatum")  # Add more species if needed, for now focus on the four species of project 1 
site_order <- c("Q16H", "Q5J", "Q5H", "M9J") # the different sand pit locations

# --------------------------
# Convert from wide to long
# --------------------------

pred_long <- prediction %>%
    dplyr::select(all_of(species_list), X, Y, site) %>%
    pivot_longer(
        cols = all_of(species_list),
        names_to = "species",
        values_to = "prediction"
    )

# Only keep sites with data
pred_long <- pred_long %>%
    filter(!is.na(prediction), site %in% site_order)

# Factorize for ordering
pred_long$species <- factor(pred_long$species, levels = species_list)
pred_long$site <- factor(pred_long$site, levels = site_order)

# --------------------------
# Color palettes per species
# --------------------------

species_palettes <- list(
    "Spisula subtruncata"    = viridis(100),
    "Abra alba"              = viridis(100),
    "Ophiura ophiura"        = viridis(100),
    "Lanice conchilega"      = viridis(100),
    "Echinocardium cordatum" = viridis(100)
)

# --------------------------
# Plotting function
# --------------------------

create_species_plot <- function(species_data, species_name){
    
    ggplot(species_data, aes(x = X, y = Y, fill = prediction)) +
        geom_raster() +
        scale_fill_gradientn(
            colors = species_palettes[[species_name]],
            name = "log(Density)",
            guide = guide_colourbar(barwidth = 1, barheight = 5)
        ) +
        facet_grid2(vars(species), vars(site), scales = "free", independent = "all") +
    theme_minimal(base_family = "Helvetica") +
        theme(
            axis.title = element_blank(),
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            strip.text.y = element_text(size = 8),
            strip.text.x = element_text(size = 8),
            panel.spacing = unit(1, "lines"),
            panel.background = element_blank(),
            panel.border = element_rect(color = "black", fill = NA, size = 1),
            legend.position = "right",
            legend.text = element_text(size = 8),
            legend.title = element_text(size = 8)
        )
}

# --------------------------
# Generate plots per species
# --------------------------

species_plots <- lapply(species_list, function(sp){
    sp_data <- pred_long %>% filter(species == sp)
    create_species_plot(sp_data, sp)
})

# Combine vertically
combined_plot <- wrap_plots(species_plots, ncol = 1)
combined_plot
```

# predict latent variables
```{r}
# only for VA models
# Obtains predictions for latent variables from a fitted generalized linear latent variable model object.
predLVs <- predictLVs.gllvm(model_Pit4, newX = new_grid, newY = NULL)
```